---
title: "Exploring livestock ownership in East Africa"
author: "Catherine Pfeifer"
date: "17 April 2018"
output:
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
rm(list = ls())

#setwd("C:/Users/Catherine Pfeifer/OneDrive/coursera/9-data product/MyFirstShinyApp")
library(shiny)
library(sp)
library(rgdal)
library(RColorBrewer)
library(dplyr)
library(ggplot2)
library(mapproj)
#library(maptools)
data<-read.csv("./dataShiny.csv")

EA.fort <- read.csv("./EAfort.csv")

liv   <- "cattle"
isrural<-1
loc<-'rural'
```

## Introduction

- DHS (Demographic Health Survey) is household data from USAID, that also contains livestock ownership information
- This is an unprecedented opportunity to undestand who keeps livestock accross the developing world
- the objective of the shiny app is to allow user to explore livestock ownership in East Africa (data for whole of Africa was too big for a free shiny app)
- the user can 
1. decide which species he/she is interested in 
2. whether he/she want to look at rural, urban or both ownerhip

The app returns a map showing the percentage of household owning the chosen type of livestock within a DHS cluster (grouped household, usually around 30 to insure that household cannot be traced back) 

## The set up of the shiny app
- the ui.r defines the interface
- the server.r contains the mapping function (more detail in the next slide)
- dataShiny.csv contains tidy DHS data 
- EA.csv contains the boundary of East Africa (Kenya, Ethiopia, Uganda). It was pre-processed with fortify (a trick to map shapefiles with ggplot)

## The mapping function 
Beyond the commands to render the in shiny, the function build on two steps

1. subset the DHS data based on the user selection with dplyr package
```{r code,eval=FALSE, echo = TRUE}
       # subset the data based on input
       if(loc=="both"){
          data2<-data%>% select(liv,x,y)%>%filter(x>0)
          } else{
          data2<-data %>% select(liv,x,y,IsRural) %>% dplyr::filter(IsRural==isrural) %>% filter(x>0)
       } # end of if loop
       data2<-na.omit(data2)
```
2. plot the data with ggplot (using the usual tricks necessary for geographical data)
```{r plot, eval=FALSE,  echo=TRUE}     
       ggplot()+
            geom_point(data=data2,aes(x,y,colour= data2[1]))+
            scale_colour_distiller(palette='Spectral')+
            geom_path(data = EA.fort,aes(long,lat,group=group))+
            coord_map()+
            ggtitle("proportion of household keeping the selected type of animals")
```

## Conclusion
1. It is a fun simple app that many stakeholders could easiely use
2. geographical data tend to bigger and therefore more heavy to run
3. do as much preprocessing outside of shiny to reduce to processing load

Enjoy exploring East Africa! 